---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r loadlibrary,  message=FALSE,echo=FALSE}
    library(dplyr)
    library(ggplot2)
```


## Loading and preprocessing the data  

1. Load the data 

```{r loaddata}
    dat <-  read.csv("activity.csv")
    str(dat)
    head(dat)
```

2. Convert variable 'date' from character type to Date type 

```{r convert}
    dat$date <- as.Date(dat$date)
    class(dat$date)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r stepsperday, message=FALSE}
    stepsperday <- dat %>%  group_by(date) %>% 
                            summarise(steps = sum(steps, na.rm = TRUE)) %>% 
                            pull(steps)
    head(stepsperday)
```

2. Histogram of the total number of steps taken each day

```{r histgram1}
   hist(stepsperday, xlab = "Number of steps per day", 
                      main = "Histogram of steps per day", col = "blue")
```

3. Mean and median of the total number of steps taken per day

```{r meanandmedian}
    mean(stepsperday, na.rm = TRUE)
    median(stepsperday, na.rm = TRUE)
```

## What is the average daily activity pattern?

1. Series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r stepsviainterval, message=FALSE}
    stepsperint <- dat %>% group_by(interval) %>% 
                            summarise(avg = mean(steps, na.rm = TRUE))
    with(stepsperint, plot(interval, avg, type = "l", col = "blue",
                             xlab = "Interval",
                             ylab = "Number of steps",
                             main = "Average number of steps via Interval"))
```

2. The interval contains the maximum number of steps

```{r findtheintervalwithmaxsteps}
    m <- stepsperint[order(-stepsperint$avg),1][1,1]
    as.numeric(m)
```

## Imputing missing values

1. Total number of missing values in the dataset

```{r missingobservations}
    sum(!complete.cases(dat))
```

2. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r filling}
    intervals <- unique(dat$interval)
    for (x in intervals) {
        dat[dat$interval == x, "steps"][is.na(dat[dat$interval == x, "steps"])] =
          mean(dat[dat$interval == x, "steps"], na.rm = TRUE)
    }

    head(dat)
```

3. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r histgram2, message=FALSE}
     sum(!complete.cases(dat))
     stepsperday2 <- dat %>% group_by(date) %>% 
                            summarise(steps = sum(steps)) %>% 
                            pull(steps)
     hist(stepsperday2, xlab = "Number of steps per day", 
                        main = "Histogram of steps per day", col = "blue")
```

```{r compare, echo=FALSE}
     m1 <- mean(stepsperday)
     m2 <- mean(stepsperday2)
     
     med1 <- median(stepsperday)
     med2 <- median(stepsperday2)
     
     print(paste0("Mean of steps per day with missing data set:", 
                  sprintf("%.2f", m1)))
     print(paste0("Mean of steps per day with filled data set:",
                  sprintf("%.2f", m2)))
     print(paste0("Median of steps per day with missing data set:",
                  sprintf("%.2f", med1)))
     print(paste0("Median of steps per day with filled data set:",
                  sprintf("%.2f", med2)))
```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r factorwday}
    dat <- dat %>% mutate(wday = if_else(weekdays(date, abbreviate = TRUE) %in%
                                           c("Sat", "Sun"), "weekend", "weekday"))
    dat$wday <- factor(dat$wday)
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r wdayandweekend, message= FALSE}
    df <- dat %>% group_by(interval, wday) %>% summarise(steps = mean(steps))
    g <- df %>% ggplot(aes(interval, steps))
    g + geom_line(col = "blue") + facet_grid(wday~.) + theme_bw() + 
      ggtitle("Average number of steps via Interval in WeekDay and Weekend")
```

